---
title: Setting up on your computer - R users
---

Here are instructions for installing on your own computer.

Install the development version of earthdatalogin and update terra.

```{r eval=FALSE}
devtools::install_github("boettiger-lab/earthdatalogin")
install.packages("terra")
install.packages("rstac")
install.packages("gdalcubes")
install.packages("here")

library("earthdatalogin")
library("terra")
library("rstac")
library("gdalcubes")
library("here")
```

You will need GDAL installed. See these instructions if you do not have it installed:  https://developers.planet.com/docs/integrations/qgis/install-qgis-gdal/

> At NOAA Fisheries Alaska Fisheries Science Center, @EmilyMarkowitz-NOAA did not need IT support to install GDAL!

You may need to install `terra` and `sf` from source to get them to use the latest GDAL installation. 

```{r eval=FALSE}
install.packages("terra", type = "source")
install.packages("sf", type = "source")
sf::sf_extSoftVersion()
```

The environment we are using today is the `py-rocket-geospatial` image. This is part of work on a [Data Science Docker Stack](https://github.com/nmfs-opensci/container-images) for NOAA Fisheries.

# R Version Metadata

```{r}
library("earthdatalogin")
library("terra")
library("rstac")
library("gdalcubes")
library("here")
library("sf")
library("dplyr")
library("stars")
library("tmap")

sessionInfo()
```
