---
title: R - ggplot2
---

::: {.callout-note icon=false}

## Goals

* Learn how to make your data frame for ggplot2
* Learn to make a x versus y plot
* Learn how to add elements to your plot
* Learn to make a boxplot
* Ask ChatGPT for help

:::

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial will give an introduction to plotting with the **ggplot2** package. Getting a basic familiarity with **ggplot2** will really save you **a lot of time**. 

This is oriented to those with little or no experience using **ggplot2** or those who have tried it and gotten frustrated. The goal is to get you started. Once you have the basics, there are a so many ggplot tutorials online -- though personally the basics are all I need.

::: {.callout-important}

If this content feels hard, remember that ChatGPT knows ggplot2. Just ask it to help you! Describe what you are trying to do.

"I have an excel file with a column called temp and a column called size. I want to read this into R and plot temp versus size using ggplot. I want red points that are shaped like triangles. The x label is temperature and the y label is size in mm. Show me only the code. "

:::

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
```


## Basic x versus y plot

### Workflow

Let's say you wanted to make a simple xy plot with `plot()`. Here's your workflow.

* Decide on `x` or use the default (1 to the number of data points).
* Decide on `y`.
* Plot with `plot(x,y)`

```{r}
val <- mtcars$mpg
x <- mtcars$hp
plot(x, val, type="p")
```


Here's your `ggplot()` workflow.

* Decide on `x` or use the default (1 to the number of data points).
* Decide on `y`.
* **Make a data frame** with `x` and `y`.
* Create the plot object with a call to `ggplot()` to tell it the data frame and then what the `x` and `y` to use: `aes(x=..., y=...)`.
* Add points or lines to the plot with `geom_line()` or `geom_point()`.

```{r}
df <- data.frame(x=mtcars$hp, val=mtcars$mpg)
p1 <- ggplot(df, aes(x=x, y=val)) + # set up
        geom_point() # plot a line
p1 # plot it
```

Typically you'd just write the call like so
```
ggplot(df, aes(x=x, y=val)) + geom_point()
```

`p1` is a ggplot object.
```{r}
class(p1)
```
We can add elements to a ggplot object
```{r}
p1 + geom_line()
```

::: {.callout-note icon=false}

### An x y plot

```
ggplot(df, aes(x=x, y=val)) +
 geom_point() +
 geom_line()
```

:::

## Modifying your plot

There are two different workflows that you need to decide on.

1. *Dynamic colors etc*: Let `ggplot()` pick your colors, points, line widths etc.
2. *Fixed colors etc*: Manually choose your colors, points, line widths etc, aka use a fixed value.

Gravitating to option 1 will make your life with `ggplot()` easier, but let's start with option 2.

::: {.callout-important}

*Fixed* lines, points attributes go *outside* of `aes()` in a `geom_...()` call. *Dynamic* attributes go *inside* of `aes()`.

:::

Look at `?geom_point` to see the attributes that you can pass in.
```{r}
ggplot(df, aes(x=x, y=val)) + geom_point(col="blue")
```

The length of the fixed attribute must be 1 or the length of the data.
```{r}
ggplot(df, aes(x=x, y=val)) + 
  geom_point(shape=21, col="blue", fill=mtcars$carb, size=6, alpha=.5, stroke=2)
```

Ways to set attributes that won't work as you think:

Putting color outside of `aes()` in `ggplot()` does nothing. `ggplot()` sets up the data to use, but information outside `aes()` doesn't flow to the plotting functions like `geom_point()`.
```{r}
ggplot(df, aes(x=x, y=val), col="blue") + geom_point()
```

Putting color inside of `aes()` in `ggplot()` has a non-intuitive effect. Plot attributes, like color, in `aes()` are converted to factors and the colors (etc) will be chosen dynamically. The name "blue" is not a color is the a factor and `ggplot()` gives the first factor the color red in this case. Information in `aes()` will *flow* to the rest of the plot unless you tell the `geom_point()` otherwise).

```{r}
ggplot(df, aes(x=x, y=val, color="blue")) + geom_point()
```

How do we *tell* `geom_point()` not to use the color value in `aes()` in `ggplot()`?  Tell it to use a fixed value by putting `col="blue"` outside of an `aes()` call in `geom_point()`.
```{r}
ggplot(df, aes(x=x, y=val)) + geom_point(col="blue")
```

What happens if we put the color in `aes()` in `geom_point()`?
```{r}
ggplot(df, aes(x=x, y=val)) + geom_point(aes(color="blue"))
```

Let's make a plot with big red points and a thick blue line.
```{r}
p1 <- ggplot(df, aes(x=x, y=val)) + 
  geom_line(col="blue", size=2) +
  geom_point(col="red", size=3)
p1
```

Let's add a column plot to that. I pass in `alpha` to add some transparency to the columns so they don't wipe out the line.
```{r}
p1 + geom_col(alpha=0.5, position="dodge")
```

### Labels and limits

With `ggplot`, use `xlab()` `ylab()` and `xlim()` `ylim()`. Ignore the NA warning. I'll show how to stop that later. 
```{r}
ggplot(df, aes(x=x, y=val)) +
  geom_point() +
  xlab("time") + ylab("value") +
  ggtitle("ggplot plot") +
  xlim(c(100,200)) + ylim(c(0,40))
```

## Changing the whole look

ggplot uses themes to set the look of your plot and you can change the whole look by setting a different theme. You can also just tweak one element of the plot's existing theme. Note because we fixed the line and point colors, we override some theme elements (eg, line colors). See `?theme_bw` to see all the themes. See `?theme` to learn how to change one element of your plot design.

```{r}
p1 + theme_classic()
```

## Adding lines or points

Let's say you want to plot 2 lines.

### Workflow

In `ggplot()` you have two possible workflows. 

Option 1. This workflow is kind of bad but often easier for beginners.

* Make a data frame for each line: `df` and `df2`.
* Call `geom_line()` with `data=df` or `data=df` and a `aes()`

```{r}
df <- data.frame(x=mtcars$hp, val=mtcars$mpg)
df2 <- data.frame(x=df$x, val2=df$val+10)
ggplot() +
  geom_point(data=df, aes(x=x, y=val)) +
  geom_line(data=df2, aes(x=x, y=val2))
```

One issue with `aes()` and data is that it is inherited rightward. This inheriting feature is usually very handy, but when working with multiple data frames, it is often clearer if you keep the data and `aes()` with the points and lines. 


Here is another example of plotting data from two different data frames:
```{r}
ggplot() +
  geom_line(data=iris, aes(x = Sepal.Length, y=Petal.Length)) +
  geom_bar(data=mtcars, aes(x=cyl), alpha=0.3) +
  ylab("y") + xlab("x")
```


### Adding a legend

ggplot makes it rather hard to modify your legend if you create a plot this way. Creating a manual legend, as opposed to dynamically as ggplot is supposed to work, can be quite hard. First thing to know is that the color, linestyle, and/or shape must be in `aes()` to appear in the legend. If it's not there you can't control it in the legend. 

Note: What I am about to show *is really hacky and not the way ggplot is intended to be used,* but it comes up so often for new ggplot users that I want you to see a solution so you don't give up on ggplot because of legends. Jump ahead to the correct ggplot workflow with long-form data frames to see how to avoid this.

This works. Ignore the warning about unknown aesthetics.
```{r}
ggplot() +
  geom_point(data=df, aes(x=x, y=val, col="df", linetype="df")) +
  geom_line(data=df2, aes(x=x, y=val2, col="df2", linetype="df2")) +
  scale_color_manual("name", values=c("blue", "black")) +
  scale_linetype_manual("name",values=c(0,1))
```

## Correct ggplot workflow

This is how `ggplot()` is intended to be used: one long data frame.

* Make data frames with `df` and `df2` data and a "name" column.
* Set up the plot with `ggplot()`
* Make points or line different using the "name" column

```{r}
df1 <- data.frame(x=x, val=val, name="df1")
df2 <- data.frame(x=x, val=val+10, name="df2")
df3 <- rbind(df1, df2)
ggplot(df3, aes(x=x, y=val, col=name, shape=name)) + 
  geom_line() +
  geom_point()
```

:::{callout-notes}

* If the color, shape, linetype is not in an `aes()` it won't appear in the legend. 
* If the color, shape, or linetype is in an `aes()` it will appear in the legend. 
* `aes()` info in `ggplot()` flows to the other elements. Put the `aes()` info in the individual `geom_...()` calls if you don't want that. Use `inherit.aes=FALSE` to stop this inheriting behavior.
* Want to mix points and lines? You need to use `scale_..._manual()` to manually turn-off points or lines for some of the data.
* You can always force colors, shapes, linetypes by passing in color, shape, size etc outside of `aes()` but it won't appear in the legend. Only colors, shapes, etc, that appear in `aes()` will appear in a legend.

Example, points are all black since the `aes(col=name)` only appears in the `geom_line()` call.
```{r}
ggplot(df3, aes(x=x, y=val)) + 
  geom_line(aes(col=name)) +
  geom_point(aes(shape=name))
```

:::

### Modifying the legend

You can control all aspects of the legend. Read up on it [here](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/).

Manually changing data points and other lines will require `scale_...()` and gets hacky.

```{r}
ggplot(df3, aes(x=x, y=val)) + 
  geom_line(aes(col=name, linetype=name)) +
  geom_point(aes(shape=name)) +
  scale_color_manual("name", values=c("blue", "black")) +
  scale_shape_manual("name",values=c(NA,1)) +
  scale_linetype_manual("name",values=c(1,0))
```

### NA warnings

Ack all those NA warnings!
```{r}
df4 <- data.frame(x=1:10, y=c(1,NA,NA,4:10), name=rep(c("a","b"),5))
ggplot(df4, aes(x=x, y=y, col=name)) + 
  geom_point()
```

Get rid of them using `na.rm=TRUE`.
```{r}
ggplot(df4, aes(x=x, y=y, col=name)) + 
  geom_point(na.rm=TRUE)
```

## Making a boxplot

Let's look at the mpg data frame.
```{r}
mpg
```

```{r}
ggplot(mpg, aes(x=drv, y=hwy)) +
  geom_boxplot()
```

```{r}
ggplot(mpg, aes(x=drv, y=hwy)) +
  geom_boxplot() +
  facet_wrap(~class) +
  xlab("") + ylab("mpg") +
  ggtitle("highway miles per gallon by class")
```


## Arranging plots into grids

## Dynamically

`ggplot()` will also dynamically break your data into plots for you.

```{r}
df1 <- data.frame(t=1:10, val=10:1, name="df1")
df2 <- data.frame(t=1:20, val=1:20+1, name="df2")
df <- rbind(df1, df2)
p1 <- ggplot(df, aes(x=t, y=val, col=name)) + geom_line()
p1
```

```{r}
p1 + facet_wrap(~name)
```

```{r}
head(mpg)
```

Let's plot city mpg versus number of cylinders.
```{r}
pc <- ggplot(mpg, aes(x=cyl, y=cty)) + geom_point()
pc
```

Now we can divide this up by different factors in our the mpg data frame.

```{r}
pc + facet_wrap(~class)
```

We can add some things to our plot and free the scales.

```{r}
pc + facet_wrap(~class, scales="free") + geom_smooth(method="lm")
```

### Manually

Now let's do this in with `ggplot()` in a similar workflow. The difference with `ggplot()` is that we save the plots and then arrange them into a grid with `gridExtra::grid.arrange()` (from the gridExtra package).

```{r warning=FALSE}
library(gridExtra)
df <- data.frame(t = 1:10, val = rnorm(10))
p1 <- ggplot(df, aes(x=t, y=val)) + geom_line() + geom_point()
p2 <- ggplot(df, aes(x=t, y=val+100)) + geom_line()
p3 <- ggplot(df, aes(x=t, y=100*val)) + geom_point()
df2 <- data.frame(t = 1:4, val = 1:4, se=.1*(1:4))
p4 <- ggplot(df2, aes(x=t, y=val)) + geom_col()
gridExtra::grid.arrange(p1, p2, p3, p4)
```

Instead of **gridExtra** and `grid.arrange()`, check out the [**patchwork**](https://cran.r-project.org/package=patchwork) package. It does similar jobs as `grid.arrange()` but is better.

Let's do two plots in one column but add new info to plot 1.

```{r}
gridExtra::grid.arrange(p1+geom_smooth(method="lm"), 
                        p4+geom_errorbar(aes(ymin=val-se, ymax=val+se), width=0.3), ncol=1)
```

When you don't know how many plots you'll be arranging. You can assemble ggplot objects into a list and pass that to `grid.arrange()`.

```{r}
df <- data.frame(t = 1:10, val = rnorm(10))
plist <- list()
n <- 3
for(i in 1:n) plist[[i]] <- ggplot(df, aes(x=t, y=val)) + geom_line() + geom_point()
df2 <- data.frame(t = 1:4, val = 1:4, se=.1*(1:4))
plist[[n+1]] <- ggplot(df2, aes(x=t, y=val)) + geom_col()
gridExtra::grid.arrange(grobs=plist)
```

The **patchwork** package gives you the same functionality.
```
patchwork::wrap_plots(plist)
```

We can combine these facet wrapped figures into a grid because they are ggplot objects. Making this plot in base R would take you forever and another 2 forevers if you wanted to change it around or if the number of classes in your data changed.

```{r}
pf <- pc + facet_wrap(~class)
grid.arrange( p1, pf, ncol=2)
```


## Notes

* `facet_wrap()` often balks if you use different data frames in your plot construction, i.e. you doing something kind of hacky.
    * needs a column in your data frame factors (or characters it can coerce into factors). Might work with multiple data frames in your plot as long as each data frame has the same "name" column.
    * wants all the data frame to be the same length. This is only when you use different data frames. Fine if you have all data in one data frame.

# Creating plot templates

If you are creating plots with the same features over and over, you can hold the features in a list and add that on to your plot.

Example where I want all my plots to have red points and a regression line:
```{r}
p3 <- geom_point(col="red")
p2 <- geom_smooth(method="lm")
p4 <- xlab("MPG")
p5 <- ylab("HP")
# pt is my template
pt <- list(p2, p3, p4, p5)

p1 <- ggplot(mtcars, aes(x=mpg, y=hp)) + pt
p2 <- ggplot(mpg, aes(x=hwy,y=cyl)) + pt
grid.arrange(p1, p2)
```

# Summary

ggplot is very powerful and will make your plotting workflow more efficient and much much faster.  It takes a little while to get the hang of, but you do not need to be a ggplot wizard. Just the basics here will take you a long way. ChatGPT knows ggplot2 and will help you.

A good set of basic ggplot commands when you are starting:

* `ggplot`
* `geom_line()`
* `geom_point()`
* `geom_col()`
* `ggtitle()`, `xlab()`, `ylab()`, `xlim()`, `ylim()`
* Themes. Use `?theme_bw` to see them.
* `gridExtra::grid.arrange(..., nrow, ncol)`
* `facet_wrap()`
* Changing the color, line, and points manually is a bit painful, but often unavoidable. Get to know the `scale_xyz_manual()` functions when you need to do that. `?scale_color_manual` to find them all.

If you want to see some examples of what ggplot2 can do. look at some galleries:

* https://r-graph-gallery.com/ggplot2-package.html
* http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html
