---
title: Geospatial in R - lab 1
---

::: {.callout-note icon=false}

## Goals

In this lab, you will learn basic skills of working with points in space.

* Read in a csv or Excel file with latitude and longitude
* Create a geospatial object using the sf package
* Plot your points with mapview and ggplot
* Apply different coordinate systems
* Tips for getting help from ChatGPT

:::

```{r message=FALSE}
require(sf)
require(mapview)
require(readr)
require(readxl)
here::i_am("r-tutorials/05-r-geospatial.qmd")
dir_data <- here::here("r-tutorials", "data")
```

In this lab, you will learn basic skills of working with points. We will store our points in data frames. All our points data frames will have these columns:

* longitude (x dimension)
* latitude (y dimension)

But the columns names are often shortened to lon, lat or lng, latd or anything else.

In addition they will have other info in other columns.

## Create a data frame of points

```{r}
library(mapview)
  
# Create example data of points
lon <- c(85.21, 80.23, 77.28)
lat = c(25.59, 12.99, 28.56)
names = c("Patna", "Chennai", "New Delhi")
  
# Create a data frame with the point data
df <- data.frame(lon, lat, names)
```
  
## Convert to a spatial points data frame

::: {.callout-important}
## Core skill

Convert data frame with latitude and longitude columns to a geospatial object with a geometry column and coordinate system. We are setting the coordinate system to [WGS 84](https://epsg.io/4326) with `crs = 4326`.

* `sf::st_as_sf()` function

:::

This is a special data frame where the location data is converted to a single point object. 

```{r}
cities3 <- sf::st_as_sf(
    df, # the data frame
    coords = c("lon", "lat"), # what are the x and y dimension names
    crs = 4326)
```

Look at the class of the object
```{r}
class(cities3)
```


## Plot the points

```{r}
plot(cities3)
```

It plotted but it is not very useful. Let's use the helper package mapview. That's more useful.

```{r}
mapview::mapview(cities3, label = cities3$names)
```

## Read in points from files

::: {.callout-important}
## Core skill

Read in tabular data with latitude, longitude into a data frame.

* `readr::read_csv()` or `readxl::read_excel()`

:::

## from a csv file

Here I use a URL to a csv file. However I could use `fil <- file.path("data", "india_tide_guages.csv")` since I have the data file in a directory `data` in the same folder as my Quarto file (or RMarkdown or R script).
```{r}
fil <- here::here("r-tutorials", "data", "india_tide_guages.csv")
df2 <- readr::read_csv(fil, show_col_types = FALSE)
```

Convert to spatial data frame. **Notice, I had to change the latitude and longitude to match the columns names in the dataframe.**

```{r}
sdf <- sf::st_as_sf(
    df2, 
    coords = c("Longitude", "Latitude"), # what are the x and y dimension names
    crs = 4326)
```

Map. You can click on the points to get more info.

```{r}
mapview::mapview(sdf)
```

If you want state labels, you need to only have the geometry and label columns in the dataframe.
```{r eval=FALSE}
sdf2 <- sdf %>% select(geometry, State)
mapview::mapview(sdf2, label = sdf2$State)
```


### from Excel file


```{r}
fil <- here::here("r-tutorials", "data", "india_tide_guages.xlsx")
df3 <- readxl::read_excel(fil, sheet = "Kerala")
```

Convert to spatial points.

```{r}
sdf = sf::st_as_sf(
    df3, 
    coords = c("Longitude", "Latitude"), 
    crs = 4326)
```

```{r}
mapview::mapview(sdf)
```

## Using ggplot2

Here is a gallery of some basic plots you can make. There are many ways to make maps with ggplot2. I will use a single approach that is fairly flexible.

::: {.callout-important}
## Core skill

Create a base world and India map from rnaturalearth. Plot with ggplot2.

* `ne_countries()`
* `ggplot() + geom_sf()`
* `coord_sf()`

:::

### India alone

```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
india_sf <- ne_countries(country = "India", scale = "medium", returnclass = "sf")
basemap <- ggplot() + 
  geom_sf(data = india_sf, color = "black", size = 2, fill="green") +
  coord_sf(xlim = c(58, 98), ylim = c(6, 30))
basemap
```

```{r}
fil <- here::here("r-tutorials", "data", "india.jpeg")
ggsave(filename = fil, plot = basemap, device = "jpeg")
```

::: {.callout-important}
## Core skill

Add points to a plot.

* `geom_sf(data=points_df)`

:::
Add points

```{r}
basemap + 
  geom_sf(data = cities3, aes(color = names), size = 3) +
  theme_void()
```

### The world

```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
world_sf <- ne_countries(scale = "medium", returnclass = "sf")
basemap <- ggplot() + 
  geom_sf(data = world_sf, color = "black", size = 0.2, fill="lightblue")
basemap
```

Add points

```{r}
basemap + 
  geom_sf(data = cities3, aes(color = names), size = 1)
```

Zoom in

```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
world_sf <- ne_countries(scale = "medium", returnclass = "sf")
basemap <- ggplot() + 
  geom_sf(data = world_sf, color = "black", size = 0.2, fill="lightblue") +
  coord_sf(xlim = c(58, 98), ylim = c(0, 30))
basemap
```

Add points

```{r}
basemap + 
  geom_sf(data = cities3, aes(color = names), size = 2) +
  coord_sf(xlim = c(58, 98), ylim = c(0, 30))
```

## Change the projection

::: {.callout-important}
## Core skill

Apply a coordinate reference system to a sf object.

* `st_transform(sf_object, crs=crs)`

Common CRS's

* `crs = 4326` [WGS 84](https://epsg.io/4326)
* Robinson `crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"`
* Globe `crs = "+proj=laea +lon_0=77 +lat_0=20 +ellps=WGS84 +no_defs"`

:::

Make a world in Robinson coord system.

```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
world_sf <- ne_countries(scale = "medium", returnclass = "sf")
crs <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" 
world_crs <- st_transform(world_sf, crs=crs)
basemap <- ggplot() +
  geom_sf(data = world_crs, color = "black", size = 0.2, fill="lightblue") +
  theme_minimal()
basemap
```

Add points.
```{r}
cities3_robin <- st_transform(cities3, crs=crs)
basemap + 
  geom_sf(data = cities3_robin, aes(color = names), size = 2)
```

Make a globe.

```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
world_sf <- ne_countries(scale = "medium", returnclass = "sf")
crs <- "+proj=laea +lon_0=77 +lat_0=20 +ellps=WGS84 +no_defs"
world_crs <- sf::st_transform(world_sf, crs)
basemap <- ggplot() +
  geom_sf(data = world_crs, color = "black", size = 0.2, fill="lightblue") +
  theme_minimal()
basemap
```

Add points.
```{r}
cities3_crs <- st_transform(cities3, crs=crs)
basemap + 
  geom_sf(data = cities3_crs, aes(color = names), size = 2)
```

Add a circle around the globe.
```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(dplyr)
world_sf <- ne_countries(scale = "medium", returnclass = "sf")
crs <- "+proj=laea +lon_0=77 +lat_0=20 +ellps=WGS84 +no_defs"
world_crs <- sf::st_transform(world_sf, crs)

sphere <- st_graticule(ndiscr = 10000, margin = 10e-6) %>%
  st_transform(crs = crs) %>%
  st_convex_hull() %>%
  summarise(geometry = st_union(geometry))

basemap <- ggplot()  +
  geom_sf(data = sphere, fill = "#D8F4FF", alpha = 0.7) +
  geom_sf(data = world_crs, fill="grey") +
  theme_bw()
basemap
```

Add points and remove legend.
```{r}
# Add crs
cities3_crs <- st_transform(cities3, crs=crs)
basemap + 
  geom_sf(data = cities3_crs, aes(color = names), size = 1) +
  theme(legend.position = "none")
```

## Getting help from ChatGPT

Unfortunately, ChatGPT often gets confused with mapping and gives you code that doesn't fully work. This can be hard for beginners (and experts) to debug.

Try telling it

* Use only the sf, rnaturalearth, and ggplot2 packages
* Work in steps, "Make a sf points object and name it sf_points", "Using sf_points", add these points to a map of the world."

## Your Turn!

Make some maps using mapview of your own data, data in the "r-tutorials/data" directory or data you can find on-line.

Try the layer feature to change the base map.


