---
title: SDM lab 1 - Downloading species data with robis
---

Here we download from OBIS using the robis package.

## Set up

```{r}
here::i_am("r-tutorials/SDM-lab-robis.qmd")
dir_data <- here::here("r-tutorials", "data")
```


```{r}
library(ggplot2)
library(sf)
library("rnaturalearth")
library("rnaturalearthdata")
library(raster)
library(tidyverse)
library(robis)
```

## Create a bounding box

```{r}
bbox <- sf::st_bbox(c(xmin = 41.875, xmax = 65.125, ymax = -0.125, ymin = 32.125), 
                          crs = sf::st_crs(4326))
```

Creates a sf object with a sfs_POLYGON from which we can get a polygon string. We also use this for cropping with the raster package, while we will need bbox for cropping with the stars package.
```{r}
extent_polygon <-  bbox %>% sf::st_as_sfc() %>% st_sf()
```

Then for the robis package we need a bounding box string.
```{r}
wkt_geometry <- extent_polygon$geometry %>% st_as_text()
```

Make a map of our region so we know we have the right area.
```{r}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot(data = world) + geom_sf() +
  geom_sf(data = extent_polygon, color = "red", fill=NA)
```

## Get occurrence data from robis

We will download data for four sea turtles found in the Arabian sea and save to one file. We will use the `occurrence()` function in the robis package.

```{r eval = FALSE}
spp <- c("Chelonia mydas", "Caretta caretta", "Eretmochelys imbricata", "Lepidochelys olivacea", "Natator depressus", "Dermochelys coriacea")

obs <- robis::occurrence(spp, startdate = as.Date("2000-01-01"), geometry = wkt_geometry)
```

This has many columns that we don't need. We reduced to fewer columns.
```{r eval = FALSE}
cols.to.use <- c("occurrenceID", "scientificName", 
                 "dateIdentified", "eventDate", 
                 "decimalLatitude", "decimalLongitude", "coordinateUncertaintyInMeters",
                 "individualCount","lifeStage", "sex",
                 "bathymetry",  "shoredistance", "sst", "sss")
obs <- obs[,cols.to.use]
```

We also added a cleaner date with YYYY-MM-DD format.
```{r eval = FALSE}
obs$date <- as.Date(obs$eventDate)
```

## Save our data

Set up the file names
```{r eval = FALSE}
dir_data <- here::here("data")
fil <- here::here("data", "io-sea-turtles.csv")
readr::write_csv(obs, file=fil)
```

Later we can reload our data as
```{r}
fil <- here::here("r-tutorials", "data", "io-sea-turtles.csv")
obs <- read.csv(fil)
```


### Clean the data

Select species.
```{r, warning=F}
# subset the occurrences to include just those in the water
obs <- obs %>% 
  subset(bathymetry > 0 & shoredistance > 0 & coordinateUncertaintyInMeters < 200)

# seeing how often each species occurs
table(obs$scientificName) 
```